Reagents: expressing and composing fine-grained concurrency

### 2011-11-01 paper/paper.tex
related work

Given our target of cache-coherent, shared-memory architectures, the
most direct way of communicating between threads is modifying shared
memory.  The challenge is to provide both \emph{atomicity} and
\emph{scalability}: communications must happen concurrently both
without corruption and without clogging the limited memory bandwidth,
even when many cores are communicating.  A simple way to provide
atomicity is to associate a lock with each shared data structure,
acquiring the lock before performing any operation.  The prevailing
wisdom\footnote{With some notable exceptions, like the recent work on
  \emph{flat combining}~\cite{?}, which we explore in \secref{?}.} is
that such \emph{coarse-grained} locking is inherently unscalable:
\begin{itemize}
\item It forces operations on the data structure to be serialized,
  even when they could be performed in parallel.
\item It adds extra cache-coherence traffic, since each core must
  acquire the same lock's cache line in exclusive mode before
  operating.  For fine-grained communication, that means at least one
  cache miss per operation.
\item It is susceptible to preemptions or stalls of a
  thread holding a lock, which prevents other threads from making progress.
\end{itemize}

@@

The design of reagents builds on a foundation laid by Concurrent ML
and extended in Haskell's STM

@@

Although CML events do not form a monad~\cite{?}, the basic pattern of
representing computations as abstract, composable values has become extremely
common in monadic programming

parallel CML's implementation of choice~\cite{Reppy2009}

@@

STM in general

@@

Communicating transactions~\cite{Lesani2011},

Transaction synchronizers~\cite{Luchangco2005},
Transaction communicators~\cite{Luchangco2011},

transactional events~\cite{Donnelly2006}

join patterns~\cite{?}

@@

@@

\newpage
\bibliographystyle{abbrvnat}
\bibliography{reagents}

@@

1.2 Title, abstract, intro,             1.0
1.0 Background                          1.0
2.5 Library overview & examples         3.0
1.5 Semantics 
1.5 Implementation/algorithms           2.0
0.5 Performance results                 1.0
1.0 Related work                        1.0
0.3 Conclusion                          0.2
0.8 Bib                                 0.8

>>>>

Given our target of cache-coherent, shared-memory architectures, the
most direct way of communicating between threads is modifying shared
memory.  The challenge is to provide both \emph{atomicity} and
\emph{scalability}: communications must happen concurrently both
without corruption and without clogging the limited memory bandwidth,
even when many cores are communicating.  A simple way to provide
atomicity is to associate a lock with each shared data structure,
acquiring the lock before performing any operation.  The prevailing
wisdom\footnote{With some dissenting opinions, as in the recent work on
  \emph{flat combining}~\cite{Hendler2010}.} is
that such \emph{coarse-grained} locking is inherently unscalable:
\begin{itemize}
\item It forces operations on the data structure to be serialized,
  even when they could be performed in parallel.
\item It adds extra cache-coherence traffic, since each core must
  acquire the same lock's cache line in exclusive mode before
  operating.  For fine-grained communication, that means at least one
  cache miss per operation.
\item It is susceptible to preemptions or stalls of a
  thread holding a lock, which prevents other threads from making progress.
\end{itemize}

@@

\elide{
The design of reagents builds on a foundation laid by Concurrent ML
and extended in Haskell's STM 
}

@@

Reagents are clearly influenced by the design of CML's events, and include
variants of CML's core event combinators (communication and choice).  Where
CML is aimed squarely at capturing synchronous communication protocols,
reagents are designed for writing and tailoring fine-grained concurrent data
structures and synchronization primitives.  This difference in motivation led
us to include a number of additional combinators, including those dealing
directly with shared state.

Originally, CML was focused on managing concurrency rather than profiting from
parallelism, and this focus was reflected in its implementation.  More
recently, a parallel implementation of CML was proposed~\cite{Reppy2009}.  The
key challenge is resolving uses of choice both consistently and scalably.  It
is addressed by \emph{sharing}: an event making a choice is enrolled as
offering a communication corresponding to each possible choice, and when a
communication is accepted, the (single, shared) event is atomically marked as
consumed.  We follow a similar strategy in dealing with message passing, but
where Parallel CML uses lock-based queues to store messages, we show how to
use lock-free bags for increased parallelism~(\secref{implementation}).  We
also show how to incorporate choice resolution with simultaneous shared-state
updates and our conjunction combinators.

\elide{
Although CML events do not form a monad~\cite{?}, the basic pattern of
representing effectful computations as abstract, composable values has become
common in monadic programming~\cite{?}, including Haskell's STM and
Transactional Events, both discussed below.
}

@@

Software transactional memory was originally intended ``to provide a general
highly concurrent method for translating sequential object implementations
into non-blocking ones''~\cite{Shavit1997}.  This ambitious goal has led to a
remarkable research literature, which has been summarized in textbook
form~\cite{Larus2006}.  Much of the research is devoted to achieving
scalability on multiprocessors or multicores, sometimes by relaxing
consistency guarantees~\cite{?} or only providing obstruction-freedom rather
than lock-freemdom~\cite{?}.  %Work in the area is

Reagents, on the other hand, are aimed at a less ambitious goal: enabling the
concise expression, user tailoring, and composition of fine-grained concurrent
algorithms.  That is, unlike STM, reagents do not attempt to provide a
\emph{universal} fine-grained concurrent algorithm.  Instead, they assist in
writing and using specific algorithms.

There is a clear tradeoff.  Using STM, one can implement a concurrent stack or
queue by simply wrapping a sequential version in an \lstinline{atomic} block,
which requires no algorithmic insight and is simpler than the stack or queue
we give in~\secref{reagents}.  But even with a very clever STM, these
implementations are unlikely to scale as well as our elimination stack or
Michael-Scott queue; some evidence for that is shown in~\secref{performance}.

Reagents carve out a middle ground between completely hand-written algorithms
and the completely automatic \lstinline{atomic} blocks of STM.  When used in
isolation, reagents are \emph{guaranteed} to perform only the CASes that the
hand-written algorithm would, so they introduce no overhead on shared-memory
operations\elide{; we know of no STM implementation that can provide similar
  guarantees or a similar cost model}.  Yet, unlike hand-written algorithms,
reagents can be tailored and composed using choice, adding blocking behavior,
or combining sequentially into larger atomic blocks. %TODO

@@

A central tenet of transactions is \emph{isolation}: transactions should not
be aware of concurrent changes to memory from other transactions.  But
sometimes it is desirable for concurrent transactions to coordinate or
otherwise communicate while executing.  Two very recent papers have proposed
mechanisms for incorporating communication with STM, in the form of
Communicating Transactions~\cite{Lesani2011} and Transaction
Communicators~\cite{Luchangco2011b} (the latter extends earlier work on
Transaction Synchronizers~\cite{Luchangco2005}).  In both cases, a key
question is how the expected isolation of shared memory can safely coexist
with concurrent communication.  Communicating Transactions use explicit,
asynchronous message passing to communicate; the mechanism is entirely
separate from shared memory, which retains isolation.  On the other hand,
Transaction Communicators allow isolation to be weakened in a controlled way.

Our mixture of message-passing and shared-state combinators most closely
resembles the work on Communicating Transactions.   %TODO

\elide{
There has also been work treating pure message-passing in a transactional way,
both explicitly in Transactional Events~\cite{Donnelly2006} and implicitly in
the Join Calculus~\cite{?}.  The former combines CML with an atomic sequencing
operator, while the latter introduces \emph{join patterns} that can receive
from multiple channels simultaneously and atomically.  Both are expressible
using reagents, through the combination of \lstinline{swap} and the
conjunction combinators.  Thus, reagents provide a new implementation for both
of these systems.  Previously, Transactional Events was implemented on top of
Haskell's STM, while join patterns have largely been implemented via
coarse-grained locking~\cite{?}.  
}

There has also been work treating pure message-passing in a transactional way:
Transactional Events~\cite{Donnelly2006} combines CML with an atomic
sequencing operator.  Previously, Transactional Events were implemented on top
of Haskell's STM, relied on lightweight search threads for matching
communications, and used an STM-based representation of channels.  However,
Transactional Events are expressible using reagents, through the combination
of \lstinline{swap} and the conjunction combinators.  Doing so yields a new
implementation that does not require search threads, performs parallel
matching for communication, and represents channels as lock-free bags.  We are
not in a position to do a head-to-head comparison, but based on the results
in~\secref{performance}, we expect the reagent-based implementation to scale
better on fine-grained workloads.

@@

Most of the literature on fine-grained concurrent data structures studies only
``within-object'' atomicity, for example developing algorithms for inserting
or removing elements into a collection atomically.  However, there has been
some nice work studying the problem of transferring data atomically between
such fine-grained data structures~\cite{Cederman2010}.  The basic approach
relies on a \emph{k}CAS operation in much the same way that reagent sequencing
does.  The transfer methods must be defined manually, in advance, and with
access to the internals of the relevant data structures, whereas reagents
allow arbitrary new compositions, without manual definition, and without
access to the code or internals of the involved data structures.

Atomic transfers between concurrent data structures

@@

\bibliographystyle{abbrvnat}
\bibliography{reagents}

@@

1.2 Title, abstract, intro,             1.0
1.0 Background                          1.0
2.5 Library overview & examples         3.0
1.5 Semantics 
1.5 Implementation/algorithms           2.0
0.5 Performance results                 1.0
1.0 Related work                        1.0
0.3 Conclusion                          0.3
0.8 Bib                                 0.7
